Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-04-03 Manuscript/manuscript.Rmd
methods edits

@@

Biological replicate Hi-C data were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1) for the GM12878 and K562 cell lines. The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from Hi-C data at 10 kb, 25 kb, 50 kb, and 100 kb resolution using the ARROWHEAD algorithm [@durand2016juicebox]. TAD boundaries called on chromosome 9 were omitted from all downstream analyses, for all resolutions and cell lines, due to the sparsity of the contact matrix at the 10 kb resolution for the K562 cell line. Genomic annotation data in the form of cell type-specific cis-regulatory elements (CRE), histone modifications (HM), and transcription factor binding sites (TFBS) were obtained from the UCSC Genome Browser Database (GRCh37/hg19 human genome assembly) [@Tyner:2017aa] (Supplementary Table 2).

@@

To make the data amenable for model training, the genome was first binned into equally sized intervals defined by the data resolution (*r*) of the corresponding contact matrix. The binning was initiated at the chromosome-specific genomic coordinate marking half of the resolution, and continued in congruent intervals until the end of the chromosome (*mod r* + *r/2*). This was to ensure that called (ground truth) TAD boundaries would be positioned in the center of genomic bins. Genomic bins were then defined as TAD boundary regions (*Y = 1*) if they contained a TAD boundary, and non-TAD boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for random forest classification.

Cell type-specific genomic annotations [???"annotations" or "elements"] were used to build the predictor space for of our random forest models. We created four types of predictors: *Peak Signal Strength* - [???include definition] (refered to hereafter as *Signal*), *Overlap Count (OC)* - the total number of overlaps between a bin and genomic annotation regions;  *Overlap Percent (OP)* - the percent of overlap between the bin and the total width of genomic annotation regions overlapping it; *Log2 Distance* - the $log_{2}$ of the distance in bases (refered to hereafter as *Distance*), from the center of the bin to the center of the nearest genomic annotation region.

The chromosome-specific dataset used for modeling was composed of two components: the response vector **Y**, an indicator variable separating between TAD and non-TAD boundary regions, and the predictor-type specific data matrix, **X** = {$X_1$, $X_2$, ..., $X_p$}, where $p$ was the set of genomic annotations. <!--One, or all types of predictors were used as described. The data was split into a 7:3 ratio of training and testing sets.-->

## Addressing class imbalance

Class imbalance (CI) was defined as the proportion of minority class out of the total number of observations:

$$CI = \dfrac{C_{min}}{C_{min} + C_{maj}}$$

where $C_{min}$ denotes the set of TAD boundary regions and $C_{maj}$ denotes the set non-TAD boundary regions. CI of heavily imbalanced classes approaches 0, while CI for balanced classes approaches 0.50. To assess the effect of CI on predictive performance, we first used the data with *No Sampling*, that is, the majority and minority classes were used as is. We evaluated three re-sampling techniques. For *Random Under-Sampling (RUS)*, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For *Random Over-Sampling (ROS)*, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For *SMOTE* (Synthetic Minority Over-Sampling Technique), under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, $n=500$. The performance of each model was compared using Matthews Correlation Coefficient (MCC), defined as <!--The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction. The MCC is defined as-->

@@

$$
MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}
$$
<!--
$$Balanced \quad Accuracy = \dfrac{1}{2}\left( \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} \right)$$
-->

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. [???include reason for using mcc, why not others?]. We refer to the complete computational framework used to compare random forest model inputs as *TADRF*. <!--Additional performance metrics including AUROC, accuracy, sensitivity, and specificity were reported in the supplementary materials.-->

## Feature reduction

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and prevent overfitting, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the best fitting *TADRF* model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity to determine the most common and top ranked annotations for predicting TAD boundaries. [??? TBD, at which stage it was done]

@@

Next, we aimed to investigate whether we could alleviate resolution limitations that plague conventional TAD calling algorithms. To do so, we extended our optimally built TAD prediction models to obtain predictions at base pair (bp) resolution. Naturally, when dealing with such fine resolution, there will exist collections of spatially close base pairs with similar predictor profiles (i.e. similar distances), and will thus yield similar predictions. To guard against inflating the number of predicted TAD boundaries we developed a technique that combines both hierarchical clustering and partitioning techniques to identify more precise TAD boundaries throughout the linear genome. We refer to our method as *preciseTAD*.

We first established a *predicted TAD boundary area* (PTBA) as a collection of adjacent base pairs that have been identified as being a predicted TAD boundary (using the optimal probability threshold identified by the combination of TPR and FPR closest to the point (0,1) on the ROC curve). We then performed hierarchical clustering on the pairwise distances (in base pairs) between the midpoints of each PTBA using complete linkage clustering. To identify the optimal number of clusters, *k*, we calculated the mean silhouette score over all possible sizes of clusters *m*:

@@

Here, $a_{i}$ is the mean distance between each cluster entry and all other observations in the same cluster, and $b_{i}$ is the mean distance between entries within the same cluster. Therefore number of clusters, *k*, that maximized the mean silhouette score, $\bar{s}_{m}$ were used to identify groupings of PTBA, which we defined as *predicted TAD boundary regions* (PTBR).

To precisely identify the PTBA most representative of each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The midpoint of the corresponding cluster medoid, in this case a particular PTBA, will be defined as a *predicted TAD boundary point* (PTBP). In the event that a particular PTBA has an even number of base pairs, then ceiling function was applied to extract the right most base pair. Thus, the hierarchical structure of our defined predicted TAD boundary objects were as follows: PTBR > PTBA > PTBP. That is, each PTBR is a collection of PTBAs, which are themselves a collection of adjacent base pairs. The uniquely identified PTBP was then the midpoint of the most representative PTBA within a PTBR.

@@

![**Assessing top ranked transcription factors in TAD boundary prediction for GM12878.** (A) Barplots comparing performances of models using specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (CRE; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all three classes (Combined; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the mean MCC across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE. Importances were calculated as the mean decrease in accuracy among permuted features in out-of-bag samples.](Figures/annots_specifc_rfe_importances.png)

>>>>

## TAD boundary data

Biological replicate Hi-C data were downloaded from GEO GSE63525 [@rao20143d] (Supplementary Table 1) for the GM12878 and K562 cell lines. The autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries were obtained from Hi-C data at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolution using the ARROWHEAD algorithm [@durand2016juicebox]. TAD boundaries called on chromosome 9 were omitted from all downstream analyses, for all resolutions and cell lines, due to the sparsity of the contact matrix at the 10 kb resolution for the K562 cell line.

@@

A total of 78 cell type-specific ChIP-Seq annotations in the form of cell cis-regulatory elements (CRE), histone modifications (HM), and transcription factor binding sites (TFBS) were obtained from the UCSC Genome Browser Database for both GM12878 and K562 (GRCh37/hg19 human genome assembly) [@Tyner:2017aa] (Supplementary Table 2).

@@

To make the data amenable for model training, the genome was first binned into equally sized intervals defined by the data resolution (*r*) of the corresponding contact matrix. The binning was initiated at the chromosome-specific genomic coordinate marking half of the resolution, and continued in congruent intervals until the end of the chromosome (*mod r* + *r/2*). This was to ensure that called (ground truth) TAD boundaries would be positioned in the center of respective genomic bins. Genomic bins were then defined as TAD boundary regions (*Y = 1*) if they contained a TAD boundary, and non-TAD boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for random forest classification.

Cell type-specific genomic annotations were used to build the predictor space for of our random forest models. We created four types of predictors: *Peak Signal Strength* - [???include definition] (refered to hereafter as *Signal*), *Overlap Count (OC)* - the total number of overlaps between a bin and genomic annotation regions;  *Overlap Percent (OP)* - the percent of overlap between the bin and the total width of genomic annotation regions overlapping it; *Log2 Distance* - the $log_{2}$ of the distance in bases (refered to hereafter as *Distance*), from the center of the bin to the center of the nearest genomic annotation region.

Chromosome-specific datasets used for modeling were composed of two components: the response vector **Y**, an indicator variable separating between TAD and non-TAD boundary regions, and the predictor-type specific data matrix, **X** = {$X_1$, $X_2$, ..., $X_p$}, where $p$ was the set of genomic annotations. Each dataset was concatenated to form a whole genome dataset. The dataset was then split into a 7:3 ratio for training and testing respectively.

To assess the impact of class imbalance (CI), defined as the proportion of TAD boundary regions to non-TAD boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Random forest classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Three-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as

@@

<!--
$$
MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}
$$
-->

$$BA = \dfrac{1}{2}\left( \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} \right)$$

@@

where TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--We refer to the complete computational framework used to compare random forest model inputs as *TADRF*.-->

## Feature reduction and predictive importance

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the feature space and prevent overfitting, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the best fitting random forest model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting TAD boundary regions.

@@

Next, we aimed to investigate whether we could alleviate resolution limitations that plague conventional TAD calling algorithms. To do so, we extended our optimally built TAD prediction models to obtain predictions at base pair (bp) resolution. Naturally, when dealing with such fine resolution, there will exist collections of spatially close base pairs with similar predictor profiles (i.e. similar distances), and will thus yield similar predictions. To guard against inflating the number of predicted TAD boundaries we developed a technique that combines both density-based spatial clustering (DBSCAN) and partitioning techniques to identify more precise TAD boundaries throughout the linear genome.

We first established a *predicted TAD boundary area* (PTBA) as a collection of adjacent base pairs that have been identified as being a predicted TAD boundary with probability 1 according to the random forest model. We then implemented DBSCAN on the distance matrix (in base pairs) constructed from the midpoints of each PTBA using the R packacge *dbscan* (version 1.1-5) [@hahsler2019dbscan]. To identify the optimal number of clusters, *k*, the eps parameter ($\epsilon$) was assigned as the respective data resolution. The heuristic being that PTBA midpoints with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3. We defined these *k* clusters of PTBAs as *predicted TAD boundary regions* (PTBR).

<!--
We then performed hierarchical clustering on the distance matrix (in base pairs) constructed from the midpoints of each PTBA using complete linkage clustering. To identify the optimal number of clusters, *k*, we calculated the mean silhouette score over all possible sizes of clusters *m*:

@@

Here, $a_{i}$ is the mean distance between each cluster entry and all other observations in the same cluster, and $b_{i}$ is the mean distance between entries within the same cluster. Therefore number of clusters, *k*, that maximized the mean silhouette score, $\bar{s}_{m}$ were used to identify groupings of PTBA, which we defined as *predicted TAD boundary regions* (PTBR).
-->

To precisely identify the PTBA most representative of each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The midpoint of the corresponding cluster medoid, in this case a particular PTBA, was defined as a *predicted TAD boundary point* (PTBP). Thus, the hierarchical structure of our defined predicted TAD boundary objects were as follows: PTBR > PTBA > PTBP. That is, each PTBR is a collection of PTBA midpoints, with the uniquely identified PTBP being the most representative midpoint within a clustered PTBR.

@@

![**Assessing top ranked transcription factors in TAD boundary prediction for GM12878.** (A) Barplots comparing performances of models using specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (CRE; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all three classes (Combined; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the mean MCC across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.](Figures/annots_specifc_rfe_importances.png)
